---
title: "Exercises 5:  Introduction to Quarto"
author:
  name: Guillem Ylla, PhD
  affiliations:
    id: JU
    name: Jagiellonian University
format: 
  html:
      embed-resources: true
      theme: lumen
      toc: true
      toc-color: red
      toc-location: left
      toc-depth: 6
      toc-expand: 6
      code-tools:
        source: true
        toggle: true
      code-fold: true
      code-summary: "Show the code"
      code-overflow: wrap
      page-layout: full
editor: visual
editor_options: 
  markdown: 
    wrap: 72
---

## Create a new Quarto document

-   In RStudio:

    -   "File" -\> "New File" -\> "Quarto Document".

        -   It will generate a new file with a short example.

-   Remember, you have 2 modes "Source" and "Visual" on the begging of
    the above menuy bar.

## Getting Started

-   Text is written in Markdown.

-   Before the Markdown, there is the YAML code that sets up some
    parameters.

    -   In this file, you will see my YAML code in which I set up the
        table of contents (toc), the code to be "folded"

-   Code Chunks by default are in R, but you can also use
    Python/Julia/Bash...

-   You can insert new code chunk with **Ctrl+Alt+i**

-   Quarto CheatSheet:
    <https://rstudio.github.io/cheatsheets/quarto.pdf>

## Run the code

-   You can run code **line by line** with **Ctrl+Enter**.

-   You can run **whole chunk** with **Ctrl+Shift+Enter**.

-   you can run **all chunks** with **Ctrl+Alt+R**.

## Render file

-   Top bar "Render"

    -   Also **Ctrl+Shift+k**.

### Exercise 1 {#sec-exercise-3 style="background-color: #fafaa0"}

1.  Add a R code chunk.
    -   In the chunk define **A** to **10** and **B** to **20**.

    -   Show values of A and B.
2.  Add another R chunk and sum A+B.
3.  Add another chunk and make a barplot of the vector containing A and
    B `c(A,B)` .
4.  Run the code:
    -   Line by line.

    -   Chunk by chunk.

    -   All chunks at once.
5.  Render.

### Solution Exercise 1 {style="background-color: #e7fccc"}

```{r}
A<-10
B<-20
```

```{r}
A+B
```

```{r}
barplot(c(A,B))
```
