---
title: "Exercises Session 7:  ggplot2"
author:
  name: Guillem Ylla, PhD
  affiliations:
    id: JU
    name: Jagiellonian University
format: 
  html:
      embed-resources: true
      theme: lumen
      toc: true
      toc-color: red
      toc-location: left
      toc-depth: 6
      toc-expand: 6
      code-tools:
        source: true
        toggle: true
      code-fold: true
      code-summary: "Show the code"
      code-overflow: wrap
      page-layout: full
editor: visual
editor_options: 
  markdown: 
    wrap: 72
---

### 

#### Exercise 1 {style="background-color: #fafaa0"}

```{r}
## YOUR CODE HERE

```

#### Solution Exercise 1 {#sec-solution-exercise-1 style="background-color: #e7fccc"}

```{r}
#| code-fold: true
#| output: true


```

```{r}
#| code-fold: true
#| output: true



```

### 

#### Exercise 2 {style="background-color: #fafaa0"}

```{r}
## YOUR CODE HERE

```

#### Solution Exercise 2 {#sec-solution-exercise style="background-color: #e7fccc"}

```{r}
#| code-fold: true
#| output: true


```

```{r}
#| code-fold: true
#| output: true



```

#### Exercise 3 {style="background-color: #fafaa0"}

```{r}
## YOUR CODE HERE

```

#### Solution Exercise 3 {style="background-color: #e7fccc"}

```{r}
#| code-fold: true
#| output: true




```

```{r}
#| eval: false
#| include: false



```

Here, we load the table that contains the Flybase ID, gene Symbol, and
Gene name:

```{r}
#| warning: false


```

#### Exercise 4 {style="background-color: #fafaa0"}

1.  Use `full_join()` to add the symbol and gene name to the long TPM
    table

```{r}
## YOUR CODE HERE

```

#### Solution Exercise 4 {style="background-color: #e7fccc"}

```{r}
#| code-fold: true
#| output: true

```

#### Exercise 5 {style="background-color: #fafaa0"}

```{r}
## YOUR CODE HERE

```

#### Solution Exercise {style="background-color: #e7fccc"}

```{r}
#| code-fold: true
#| output: true


```
