---
title: "Quarto Notebook #2"
author: "Guillem Ylla"
format: 
  html:
    toc: true
    theme: lumen
    code-fold: FALSE
editor: visual
execute:
  cache: false
---

# Guillem Lab Notebook #2

-   Project: Teaching Quarto #2

-   Date: 04/01/2024

## YAML code (1st Chunk)

Allows us to set different parameters to the Quarto notebook and how it will be rendered.

## Chunk options

### Old way

In the top right corner of the chunk we have different options (in source mode) to decide what to do with the chunk. This method comes from the Rmarkdwon times.

The following chunk will show code but don't run it because include=TRUE and eval=FALSE in the chunk header

```{r eval=FALSE, include=TRUE}
1 + 1
```

### Modern way

In the YAML, by default we defined "code-fold: FALSE". But now, I want to fold following chunk. We can modify the behavior adding "#\| code-fold: TRUE" in the top of the chunk.

```{r}
#| label: chunk-two-two
#| code-fold: TRUE	

2+2
```

Now we will see the code ("#\| echo: true") but won't run it ("#\| eval: false")

```{r}
#| label: chunk-tree-three
#| echo: true
#| eval: false

3+3
```

```{r}
#| label: Chunk-declare_vars
#| cache: true

A=5
B=5
C=10
```

```{r}
#| label: sum

A+B
```

## Add figure

```{r}
#| label: fig-figure
#| fig-cap: Caption for the figure 

barplot(c(A,B,C))

```

## Cross-reference

I can here reference to any element such as a figure @fig-figure (chunk label must start with *fig-*) or tables. Using @ or Insert-\>Cross Reference .

## Bibliography

## Session Info

```{r}
sessionInfo()
```
